---
title: "Comparison of Prediction Market Data vs. Economist Model"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(googlesheets4)
library(googledrive)
library(tidyverse)
library(conflicted)
library(slider)
library(scales)
library(assertthat)
library(here)
library(patchwork)
library(ggthemes)
library(readxl)
library(tools)
conflicted::conflict_prefer_all("dplyr", quiet = TRUE)
#googledrive::drive_auth()

state_legend = theme(legend.position = "right")
no_state_legend = theme(legend.position = "none")

plot_theme <- theme_minimal(base_size = 12) +
  theme(
    text = element_text(family = "Helvetica", color = "#222222"),
    plot.title = element_text(size = 14, face = "bold", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 11, margin = margin(b = 10)),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 9, color = "#333333"),
    panel.grid.major = element_line(color = "gray90", size = 0.2),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.35, "cm"),
    legend.spacing.x = unit(0.1, "cm")
  )

```

# 2020 Election


## Market Data

```{r}

mkt_2020 <- read_xlsx("/Users/ebbamark/OneDrive - Nexus365/Market_Data/Price History By Market -Bulk-Presidential.xlsx") %>% 
  mutate(state = gsub(" in the  2020 presidential election?", "", gsub(" in the 2020 presidential election?", "", gsub("Which party will win ", "", MarketName), fixed = TRUE), fixed = TRUE))

# mkt_2020 %>% 
#   summarise(across(everything(), ~n_distinct(.)))

stopifnot(mkt_2020 %>% 
  group_by(MarketId, HistoryDateET, ContractName) %>% 
  distinct %>% 
  n_groups(.) == nrow(mkt_2020))

mkt_2020 %>% 
  group_by(HistoryDateET, ContractName) %>% 
  summarise(CloseSharePrice = mean(CloseSharePrice)) %>%
  filter(ContractName == "Democratic") %>% 
  ggplot(aes(x = HistoryDateET, y = CloseSharePrice)) +
  geom_line(color = "#440154FF") +
  theme_fivethirtyeight() +
  labs(y = "Daily Closing Share Price", 
       x = "Date", 
       color = "Party", 
       title = "Daily Closing Share Price on Democratic Contracts for\n2020 US Presidential Election",
       subtitle = "By definition, the closing share price of Republican contracts is $1-closing share price \nof Democratic Contracts") +
    plot_theme


mkt_2020 %>% 
  filter(ContractName == "Democratic") %>% 
  ggplot(aes(x = HistoryDateET, color = state, fill = state)) +
  geom_ribbon(aes(ymin = LowSharePrice, ymax = HighSharePrice), alpha = 0.1) +
  geom_line(aes(y = CloseSharePrice)) +
  theme_fivethirtyeight() +
  labs(y = "Daily Closing Share Price", 
       x = "Date", 
       
       title = "Daily Closing Share Price on Democratic Contracts for\n2020 US Presidential Election by State",
       subtitle = "By definition, the closing share price of Republican contracts is $1-closing share price \nof Democratic Contracts") +
    plot_theme +
  no_state_legend +
      scale_color_viridis_d(option = "D", direction = 1)


```



## Economist Data
```{r}

rec_string <- 'electoral-college-probability-over-time.xlsx'

files <- list.files("/Users/ebbamark/OneDrive - Nexus365/Economist_Data") %>% .[grepl(rec_string, .)]

econ_2020 <- tibble()
for(file in files){
  state_name <- sub("-electoral-college-probability-over-time\\.xlsx$", "", file) %>% 
    gsub("-"," ", .) %>% 
    toTitleCase(.)
  econ_2020 <- read_xlsx(paste0("/Users/ebbamark/OneDrive - Nexus365/Economist_Data/", file)) %>% 
    mutate(state = state_name) %>% rbind(econ_2020,.)
}

stopifnot(econ_2020 %>% 
  group_by(date, state) %>% 
  n_groups(.) == nrow(econ_2020)/2)

stopifnot(econ_2020 %>% 
  group_by(date, state) %>% 
  summarise(win_prob = round(sum(win_prob))) %>% 
  filter(win_prob != 1) %>% nrow(.) == 0)

econ_2020 %>% 
  filter(party == "democratic") %>% 
  #group_by(date) %>% 
  #summarise(mean_win_prob = mean(win_prob)) %>% 
  ggplot(aes(x = date, y = win_prob, color = state)) +
  geom_line() +
  theme_fivethirtyeight() +
  labs(x ="Date", y = "Democratic Win Probability", title = "Economist - Democratic Win Probability in \n2020 Presidential Election by State") +
  plot_theme +
  no_state_legend +
      scale_color_viridis_d(option = "D", direction = 1) +
        scale_fill_viridis_d(option = "D", direction = 1)



```


```{r}

# Produces plots to compare the market price and economist prediction

comp_plots <- function(df, year, vir_pal = "D"){
  
  if(year == "2020"){
      shading_color <- "slateblue"
      df$state <- state.abb[match(df$state, state.name)]
  }else{
    shading_color <- "mediumvioletred"
  }
  
  box_plot <- df %>% 
    mutate(state = reorder(state, diff, FUN = mean)) %>%
    ggplot() +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.00, ymax = 0.1, fill = shading_color, alpha = 0.15) +
  #annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.05, ymax = 0.1, fill = shading_color, alpha = 0.2) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.10, ymax = 0.2, fill = shading_color, alpha = 0.1) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.2, ymax = 0.3, fill = shading_color, alpha = 0.05) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.3, ymax = 0.4, fill = shading_color, alpha = 0.01) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.00, ymax = -0.1, fill = shading_color, alpha = 0.15) +
  #annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.05, ymax = 0.1, fill = shading_color, alpha = 0.2) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = -0.10, ymax = -0.2, fill = shading_color, alpha = 0.1) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = -0.2, ymax = -0.3, fill = shading_color, alpha = 0.05) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = -0.3, ymax = -0.4, fill = shading_color, alpha = 0.01) +

    geom_boxplot(aes(x = reorder(state, diff, FUN = mean), y = diff, color = state, fill = state), alpha = 0.1) +
    plot_theme +
    no_state_legend +
    scale_color_viridis_d(option = vir_pal, direction = 1, begin = 0, end = 0.85) +
    scale_fill_viridis_d(option = vir_pal, direction = 1, begin = 0, end = 0.85) +
    labs(x = "State", y = "Closing Share Price - Economist Forecast", title = paste0(year, " Election: Absolute difference between Closing Market Price and Economist Projection by State")) +
    scale_y_continuous(breaks = seq(-1, 1, by = 0.1))
  
  # MSE by state (bar chart) 
  mse_by_state <- df %>%
    group_by(state) %>%
    summarise(
      n_obs = sum(!is.na(sq)),
      mse = mean(sq, na.rm = TRUE)
    ) %>%
    ungroup() %>%
    arrange(desc(mse))
  
  p_bar <- mse_by_state %>%
    mutate(state = reorder(state, mse, FUN = max)) %>%
    ggplot(aes(x = state)) +
    geom_point(aes(y = mse)) +
    geom_segment(aes(y = mse, yend = 0, color = state)) +
    #coord_flip() +
    labs(
      title = paste0("Mean Squared Error by State (", year,")"),
      x = NULL,
      y = "MSE"
    ) +
        plot_theme +
    no_state_legend +
    scale_color_viridis_d(option = vir_pal, direction = -1)
  
  
  if(year == 2020){
  # Rolling MSE over time (facet per state) 
  # rolling window (days) - change window_days as desired
  window_days <- 7
  
  # ensure data is sorted by state/date
  df_sorted <- df %>%
    arrange(state, date)
  
  # compute rolling mean of squared error per state
  rolling_df <- df_sorted %>%
    group_by(state) %>%
    mutate(
      mse_roll = slide_dbl(sq, ~ mean(.x, na.rm = TRUE), .before = window_days - 1, .complete = TRUE)
    ) %>%
    ungroup()
  
top_n <- 12
top_states <- mse_by_state %>%
  arrange(desc(mse)) %>%
  slice_head(n = top_n) %>%
  pull(state)

# ---- plot A: top 12 states in grid ----
p_top12 <- rolling_df %>%
  filter(state %in% top_states) %>%
  ggplot(aes(x = date, y = mse_roll, color = state)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  facet_wrap(~ state, scales = "free_y", ncol = 4) +
  labs(
    title = paste0(window_days, "-Day Rolling MSE — Top ", top_n, " States"),
    subtitle = "Each facet shows the rolling mean squared error; scales are free per panel",
    x = "Date",
    y = "Rolling MSE"
  ) +
  scale_color_viridis_d(option = vir_pal, direction = -1) +
  plot_theme +
  theme(legend.position = "none")

# ---- plot B: remaining states together ----
p_rest <- rolling_df %>%
  filter(!(state %in% top_states)) %>%
  ggplot(aes(x = date, y = mse_roll, group = state)) +
  geom_line(color = "gray60", linewidth = 0.4, alpha = 0.6) +
  labs(
    title = "Remaining States (combined)",
    subtitle = "All other states shown together for context",
    x = "Date",
    y = "Rolling MSE"
  ) +
  plot_theme

# ---- combine vertically using patchwork ----
final_plot <- p_top12 / p_rest +
  plot_layout(heights = c(3, 1))  # top grid larger, bottom smaller
  }
  
  
  print(p_bar)
  print(box_plot)
  
  if(year == "2020"){
  print(final_plot)}
  
  return(list(p_bar, box_plot))

}

```

## Comparing State by State

In 2020, the Economist reports on 17 states whereas the prediction market reports on 56 states/regions (Maine and Nebraska in 3 separate regions each).

```{r}

# setdiff(econ_2020$state, mkt_2020$state)
# setdiff(mkt_2020$state, econ_2020$state)
# 
# setdiff(econ_2020$date, mkt_2020$HistoryDateET)
# setdiff(mkt_2020$HistoryDateET, econ_2020$date)

mkt_2020_comp <- mkt_2020 %>% 
  select(ContractName, HistoryDateET, CloseSharePrice, state) %>% 
  filter(ContractName == "Democratic") %>% 
  rename(date = HistoryDateET) %>% 
  select(date, state, CloseSharePrice)

econ_2020_comp <- econ_2020 %>% 
  filter(party == "democratic") %>% 
  select(date, state, win_prob)

df_comp_2020 <- mkt_2020_comp %>%
  left_join(econ_2020_comp, by = c("date", "state")) %>%
  filter(!is.na(CloseSharePrice) & !is.na(win_prob)) %>%
  mutate(diff = CloseSharePrice - win_prob,
         sq = diff^2)

plots_2020 <- comp_plots(df_comp_2020, year = "2020")



```


# 2016 Election

## Market Data

```{r}

mkt_2016 <- read_xlsx("/Users/ebbamark/OneDrive - Nexus365/Market_Data/Price History By Market 2016.xlsx") %>% 
  mutate(state = gsub("the ", "", gsub(" in the 2016 presidential election?", "", gsub("Which party will win ", "", MarketName), fixed = TRUE)))

stopifnot(mkt_2016 %>% group_by(state, HistoryDateET, ContractName) %>% n_groups(.) == nrow(mkt_2016))


stopifnot(mkt_2016 %>% 
  group_by(MarketId, HistoryDateET, ContractName) %>% 
  distinct %>% 
  n_groups(.) == nrow(mkt_2016))

mkt_2016 %>% 
  group_by(HistoryDateET, ContractName) %>% 
  summarise(CloseSharePrice = mean(CloseSharePrice)) %>%
  filter(ContractName == "Democratic") %>% 
  ggplot(aes(x = HistoryDateET, y = CloseSharePrice)) +
  geom_line(color = "#B12A90FF") +
  theme_fivethirtyeight() +
  labs(y = "Daily Closing Share Price", 
       x = "Date", 
       color = "Party", 
       title = "Daily Closing Share Price on Democratic Contracts for\n2016 US Presidential Election",
       subtitle = "By definition, the closing share price of Republican contracts is $1-closing share price \nof Democratic Contracts") +
  plot_theme


mkt_2016 %>% 
  filter(ContractName == "Democratic") %>% 
  ggplot(aes(x = HistoryDateET, color = state, fill = state)) +
  geom_ribbon(aes(ymin = LowSharePrice, ymax = HighSharePrice), alpha = 0.1) +
  geom_line(aes(y = CloseSharePrice)) +
  theme_fivethirtyeight() +
  labs(y = "Daily Closing Share Price", 
       x = "Date", 
       title = "Daily Closing Share Price on Democratic Contracts for\n2016 US Presidential Election",
       subtitle = "By definition, the closing share price of Republican contracts is $1-closing share price \nof Democratic Contracts") +
    plot_theme +
  no_state_legend +
        scale_color_viridis_d(option = "C", direction = 1)

```

## Economist Data

```{r}

econ_2016 <- read_xlsx("/Users/ebbamark/OneDrive - Nexus365/Economist_Data/loop_summaries_2016.xlsx")

stopifnot(econ_2016 %>% group_by(state, model_run_date) %>% n_groups(.) == nrow(econ_2016))

econ_2016 %>% 
  filter(!(state %in% c("--", "ELECTORAL COLLEGE"))) %>% 
  ggplot(aes(x = model_run_date, y = mean, color = state)) +
  geom_ribbon(aes(x = model_run_date, ymin = low, ymax = high, fill = state, color = NA), alpha = 0.2) +
  geom_line() +
  theme_fivethirtyeight() +
  labs(x = "Model Run Date", y = "Mean Win Probability", title = "Economist - Democratic Win Probability in \n2016 Presidential Election") +
    plot_theme +
  no_state_legend +
        scale_color_viridis_d(option = "C", direction = 1) +
          scale_fill_viridis_d(option = "C", direction = 1)


```

## Comparing State by State

In 2016, the Economist reports on 51 states whereas the prediction market reports on 51 states/regions (both include DC).

```{r}

mkt_2016_comp <- mkt_2016 %>% 
  select(ContractName, HistoryDateET, CloseSharePrice, state) %>% 
  filter(ContractName == "Democratic") %>% 
  rename(date = HistoryDateET) %>% 
  select(date, state, CloseSharePrice)

mkt_2016_comp$state_name <- state.abb[match(mkt_2016_comp$state, state.name)]

mkt_2016_comp <- mkt_2016_comp %>% 
  mutate(state_name = case_when(state == "Washington state" ~ "WA",
                                state == "District of Columbia" ~ "DC",
                                TRUE ~ state_name)) %>% 
  select(-state) %>% 
  rename(state = state_name)


econ_2016_comp <- econ_2016 %>% 
  filter(!(state %in% c("--", "ELECTORAL COLLEGE"))) %>% 
  rename(date = model_run_date,
         win_prob = mean) %>% 
  select(date, state, win_prob)

df_comp_2016 <- mkt_2016_comp %>%
  left_join(econ_2016_comp, by = c("date", "state")) %>%
  filter(!is.na(CloseSharePrice) & !is.na(win_prob)) %>%
  mutate(diff = CloseSharePrice - win_prob,
         sq = diff^2)

plots_2016 <- comp_plots(df_comp_2016, vir_pal = "C", year = "2016")

plots_2016[[2]] / plots_2020[[2]]

ggsave(here("data/absolute_difference.png"), height = 10, width = 13)


```

```{r}


# MSE by state (bar chart) 
mse_by_state_2016 <- df_comp_2016 %>%
    group_by(state) %>%
    summarise(
      n_obs = sum(!is.na(sq)),
      mse = mean(sq, na.rm = TRUE),
      rmse = sqrt(mse)
    ) %>%
    ungroup() %>%
    arrange(desc(mse)) %>% 
  mutate(year = 2016)

mse_by_state_2020 <- df_comp_2020 %>% 
    group_by(state) %>%
    summarise(
      n_obs = sum(!is.na(sq)),
      mse = mean(sq, na.rm = TRUE),
      rmse = sqrt(mse)
    ) %>%
    ungroup() %>%
    arrange(desc(mse)) %>% 
  mutate(year = 2020)

mse_by_state_2020$state <- state.abb[match(mse_by_state_2020$state, state.name)]


mse_by_state_2016 %>% 
  full_join(., mse_by_state_2020, by = c("state"), suffix = c("_2016", "_2020")) %>% 
  mutate(state = reorder(state, rmse_2016, FUN = max)) %>%
    ggplot(aes(x = state)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.00, ymax = 0.1, fill = "mediumseagreen", alpha = 0.2) +
  #annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.05, ymax = 0.1, fill = "mediumseagreen", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.10, ymax = 0.2, fill = "mediumseagreen", alpha = 0.1) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.2, ymax = 0.3, fill = "mediumseagreen", alpha = 0.05) +
      #annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.3, ymax = 0.4, fill = "mediumseagreen", alpha = 0) +

    geom_point(aes(y = rmse_2016, color = "2016")) +
    geom_point(aes(y = rmse_2020, color = "2020")) + 
    geom_ribbon(aes(x = state, ymin = 0, ymax=0.1), fill = "grey70") +
    labs(
      title = "Gap between Prediction Market Price and Economist Forecast of State Presidential Election Returns",
      subtitle = "Statistic: Root Mean Squared Error by State in 2016 (May 10 - Nov. 8) and 2020 (Mar. 1 - Nov. 3)",
      x = "State",
      y = "RMSE"
    ) +
  #coord_flip() +
    plot_theme +
    theme(legend.position = "bottom") +
  scale_color_viridis_d(option = "viridis", begin = 0.6, end = 0.2) 


ggsave(here("data/market_price_model_divergence.png"), height = 9, width = 12)


```


